The role of cognateness in native spoken word recognition

### 2025-07-24 docs/manuscript.qmd
Update

```{r}
#| label: setup
library(targets)
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
library(gt)
library(gtExtras)
library(brms)
library(tidybayes)
library(ggExtra)
library(tinytable)
library(stringdist)
set.seed(1234)

# import objects
tar_config_set(
  store = here::here("_targets"),
  script = here::here("_targets.R")
)
tar_load_globals()
tar_load(participants)
tar_load(exp_participants)
tar_load(exp_responses)
tar_load(quest_responses)
tar_load(quest_participants)
tar_load(stimuli)
tar_load(epreds)
tar_load(dataset_1)
tar_load(dataset_2)
tar_load(dataset_3)
tar_load(exp_1_m0)
tar_load(exp_2_m0)
tar_load(exp_3_m1)
tar_load(exp_12_m0)

theme_set(theme_ggdist() +
  theme(panel.grid.major.y = element_line(
    colour = "grey",
    linetype = "dotted"
  )))

@@

```{r tbl-stimuli}
#| label: tbl-stimuli
#| tbl-cap: "Stimuli details."
stimuli |>
  summarise(
    across(c(freq_zipf_2, lv, neigh_n_h),
      lst(mean, sd, min, max),
      na.rm = TRUE
    ),
    .by = c(group)
  ) |>
  gt(rowname_col = "group") |>
  tab_spanner("Frequency", matches("freq")) |>
  tab_spanner("Similarity", matches("lv")) |>
  tab_spanner("CLPN", matches("neigh")) |>
  fmt_number(is.numeric) |>
  fmt_integer(c(
    neigh_n_h_min,
    neigh_n_h_max
  )) |>
  cols_merge_uncert(freq_zipf_2_mean, freq_zipf_2_sd) |>
  cols_merge_uncert(lv_mean, lv_sd) |>
  cols_merge_uncert(neigh_n_h_mean, neigh_n_h_sd) |>
  cols_merge_range(freq_zipf_2_min, freq_zipf_2_max) |>
  cols_merge_range(lv_min, lv_max) |>
  cols_merge_range(neigh_n_h_min, neigh_n_h_max) |>
  cols_label(
    freq_zipf_2_mean = "Mean ± SD",
    freq_zipf_2_sd = "SD",
    freq_zipf_2_min = "Range",
    lv_mean = "Mean ± SD",
    lv_sd = "SD",
    lv_min = "Range",
    neigh_n_h_mean = "Mean ± SD",
    neigh_n_h_sd = "SD",
    neigh_n_h_min = "Range"
  ) |>
  tab_style(
    cell_text(weight = "bold"),
    cells_column_spanners()
  ) |>
  tab_style(
    cell_text(style = "italic"),
    cells_column_labels()
  ) |>
  grand_summary_rows(
    columns = matches("mean|sd"),
    fns = list(
      label = md("**Mean**"),
      id = "totals",
      fn = "mean"
    ),
    fmt = ~ fmt_number(., n_sigfig = 3, suffixing = TRUE)
  )
```

@@

All analyses were performed in the R environment [@rcoreteam2013language]. We used the tidyverse family of R packages [@wickham2019welcome] to process data and to generate figures. We used the `brms` R package [@burkner2017brms] using the `cmdstanr` back-end to the Stan probabilistic language [@carpenter2017stan] to estimate and compare the models (see Appendix 1 for model diagnostics). Numeric predictors were standardised before entering the model by subtracting the mean and dividing by the standard deviation.

@@

```{r tbl-dataset}
#| label: tbl-dataset
#| tbl-cap: "Summary of participants' accuracy in the translation elicitation task across Experiments 1, 2, and 3."
sem <- function(x) mean(x) / (sqrt(length(x)))

@@

```{r fig-knowledge}
#| label: fig-knowledge
#| fig-width: 9
#| fig-height: 4
#| fig-cap: "Catalan and Spanish prior word knowledge reported by English native participants in Experiment 3. The X-axis indicates the proportion of participants that reported prior knowledge with the word. The Y-axis indicates participants' accuracy (the proportion of participants that translated the word correctly). For visualisation purposes, data points have been aggregated so that numbers and color coding show the number of data points with identical accuracy and proportion of reported knowledge."

bin_interval <- seq(0, 1, 0.05)
dataset_3 |>
  mutate(surprise = translation %in% tbl_data$Translation) |>
  summarise(
    knowledge = mean(knowledge),
    correct = sum(correct),
    n = n(),
    .by = c(translation, group, surprise)
  ) |>
  mutate(
    accuracy = prop_adj(correct, n),
  accuracy_binned = cut(accuracy, bin_interval, labels = bin_interval[-length(bin_interval)], include.lowest = TRUE, right = TRUE),
    knowledge_binned = cut(knowledge, bin_interval, labels = as.numeric(bin_interval[-length(bin_interval)]), include.lowest = TRUE, right = TRUE)
  ) |> 
    summarise(
      n = n(),
      .by = c(accuracy_binned, knowledge_binned, group)
    ) |> 
      mutate(is_gt_20 = n > 20,
      knowledge_binned = as.double(as.character(knowledge_binned)),
      accuracy_binned = as.double(as.character(accuracy_binned))) |> 
  ggplot(aes(knowledge_binned, accuracy_binned, fill = n, label = n, color = is_gt_20)) +
  facet_wrap(~group) +
  geom_tile(color = NA) +
  geom_text(size = 3) +
  labs(
    x = "Reported knowledge",
    y = "Accuracy"
  ) +
  scale_fill_distiller(palette = "Reds", direction=1, limits = c(0, 37)) +
  scale_color_manual(values = c("black", "white")) +
  scale_x_continuous(labels = scales::percent, limits = c(-0.05, 1)) +
  scale_y_continuous(labels = scales::percent, limits = c(-0.05, 1), breaks = seq(0, 1, 0.1)) +
  theme(
    legend.position = "none",
  )
```

@@

```{r fig-epreds-3}
#| label: fig-epreds-3
#| cache: false
#| fig-cap: "Posterior model-predicted mean accuracy in Experiment 3. Predictions were generated from 4,000 posterior samples, extracted for different values of *CLPN* (0, 2, 4, 8, 12) and *Similarity* (1-100). Predictions are plotted separately for English participants translating Catalan words, and for English participants translating Spanish words. Lines indicate mean predictions, and intervals indicate 95%, 89%, 78%, 67%, and 50% credible intervals (CrI)."
#| fig-height: 5
#| fig-width: 10
lv <- seq(0, 1, length.out = 100)
lv_std <- (lv - mean(dataset_3$lv_std, na.rm = TRUE)) / sd(dataset_3$lv_std, na.rm = TRUE)
neigh_n_h <- c(0, 2, 4, 8, 12)
epreds_3 <- get_epreds(exp_3_m1, dataset_3, neigh_n_h = neigh_n_h, lv = lv) 
epreds_3$neigh_n_h_std <- factor(epreds_3$neigh_n_h_std, levels = unique(epreds_3$neigh_n_h_std), labels = paste0(neigh_n_h, " neighbours"), ordered = TRUE)

@@

# References

>>>>

```{r}
#| label: setup
library(targets)
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
library(gt)
library(gtExtras)
library(brms)
library(tidybayes)
library(ggExtra)
library(tinytable)
library(stringdist)

set.seed(1234)

# import objects
tar_config_set(
  store = here::here("_targets"),
  script = here::here("_targets.R")
)

tar_load_globals()
tar_load(participants)
tar_load(exp_participants)
tar_load(exp_responses)
tar_load(quest_responses)
tar_load(quest_participants)
tar_load(stimuli)
tar_load(epreds)
tar_load(dataset_1)
tar_load(dataset_2)
tar_load(dataset_3)
tar_load(exp_1_m0)
tar_load(exp_2_m0)
tar_load(exp_3_m1)
tar_load(exp_12_m0)

theme_set(theme_ggdist() +
  theme(panel.grid.major.y = element_line(
    colour = "grey",
    linetype = "dotted"
  )))
```

@@

```{r}
#| label: tbl-stimuli
#| tbl-cap: "Stimuli details."
stimuli |>
  summarise(
    across(c(freq_zipf_2, lv, neigh_n_h),
      lst(mean, sd, min, max),
      na.rm = TRUE
    ),
    .by = c(group)
  ) |>
  gt(rowname_col = "group") |>
  tab_spanner("Frequency", matches("freq")) |>
  tab_spanner("Similarity", matches("lv")) |>
  tab_spanner("CLPN", matches("neigh")) |>
  fmt_number(is.numeric) |>
  fmt_integer(c(
    neigh_n_h_min,
    neigh_n_h_max
  )) |>
  cols_merge_uncert(freq_zipf_2_mean, freq_zipf_2_sd) |>
  cols_merge_uncert(lv_mean, lv_sd) |>
  cols_merge_uncert(neigh_n_h_mean, neigh_n_h_sd) |>
  cols_merge_range(freq_zipf_2_min, freq_zipf_2_max) |>
  cols_merge_range(lv_min, lv_max) |>
  cols_merge_range(neigh_n_h_min, neigh_n_h_max) |>
  cols_label(
    freq_zipf_2_mean = "Mean ± SD",
    freq_zipf_2_sd = "SD",
    freq_zipf_2_min = "Range",
    lv_mean = "Mean ± SD",
    lv_sd = "SD",
    lv_min = "Range",
    neigh_n_h_mean = "Mean ± SD",
    neigh_n_h_sd = "SD",
    neigh_n_h_min = "Range"
  ) |>
  tab_style(
    cell_text(weight = "bold"),
    cells_column_spanners()
  ) |>
  tab_style(
    cell_text(style = "italic"),
    cells_column_labels()
  ) |>
  grand_summary_rows(
    columns = matches("mean|sd"),
    fns = list(
      label = md("**Mean**"),
      id = "totals",
      fn = "mean"
    ),
    fmt = ~ fmt_number(., n_sigfig = 3, suffixing = TRUE)
  )
```

@@

All analyses were performed in the R environment [@rcoreteam2013language]. We used the tidyverse family of R packages [@wickham2019welcome] to process data and to generate figures. We used the `brms` R package [@burkner2017brms] using the `cmdstanr` back-end to the Stan probabilistic language [@carpenter2017stan] to estimate and compare the models (see #apx-diagnostics for model diagnostics). Numeric predictors were standardised before entering the model by subtracting the mean and dividing by the standard deviation.

@@

```{r}
#| label: tbl-dataset
#| tbl-cap: "Summary of participants' accuracy in the translation elicitation task across Experiments 1, 2, and 3."
sem <- function(x) mean(x) / (sqrt(length(x)))

@@

```{r}
#| label: fig-knowledge
#| fig-width: 9
#| fig-height: 4
#| fig-cap: "Catalan and Spanish prior word knowledge reported by English native participants in Experiment 3. The X-axis indicates the proportion of participants that reported prior knowledge with the word. The Y-axis indicates participants' accuracy (the proportion of participants that translated the word correctly). For visualisation purposes, data points have been aggregated so that numbers and color coding show the number of data points with identical accuracy and proportion of reported knowledge."
bin_interval <- seq(0, 1, 0.05)
dataset_3 |>
  mutate(surprise = translation %in% tbl_data$Translation) |>
  summarise(
    knowledge = mean(knowledge),
    correct = sum(correct),
    n = n(),
    .by = c(translation, group, surprise)
  ) |>
  mutate(
    accuracy = prop_adj(correct, n),
  accuracy_binned = cut(accuracy, bin_interval, labels = bin_interval[-length(bin_interval)], include.lowest = TRUE, right = TRUE),
    knowledge_binned = cut(knowledge, bin_interval, labels = as.numeric(bin_interval[-length(bin_interval)]), include.lowest = TRUE, right = TRUE)
  ) |> 
    summarise(
      n = n(),
      .by = c(accuracy_binned, knowledge_binned, group)
    ) |> 
      mutate(is_gt_20 = n > 20,
      knowledge_binned = as.double(as.character(knowledge_binned)),
      accuracy_binned = as.double(as.character(accuracy_binned))) |> 
  ggplot(aes(knowledge_binned, accuracy_binned, fill = n, label = n, color = is_gt_20)) +
  facet_wrap(~group) +
  geom_tile(color = NA) +
  geom_text(size = 3) +
  labs(
    x = "Reported knowledge",
    y = "Accuracy"
  ) +
  scale_fill_distiller(palette = "Reds", direction=1, limits = c(0, 37)) +
  scale_color_manual(values = c("black", "white")) +
  scale_x_continuous(labels = scales::percent, limits = c(-0.05, 1)) +
  scale_y_continuous(labels = scales::percent, limits = c(-0.05, 1), breaks = seq(0, 1, 0.1)) +
  theme(
    legend.position = "none",
  )
```

@@

```{r}
#| label: fig-epreds-3
#| cache: false
#| fig-cap: "Posterior model-predicted mean accuracy in Experiment 3. Predictions were generated from 4,000 posterior samples, extracted for different values of *CLPN* (0, 2, 4, 8, 12) and *Similarity* (1-100). Predictions are plotted separately for English participants translating Catalan words, and for English participants translating Spanish words. Lines indicate mean predictions, and intervals indicate 95%, 89%, 78%, 67%, and 50% credible intervals (CrI)."
#| fig-height: 5
#| fig-width: 10
lv <- seq(0, 1, length.out = 100)
lv_std <- (lv - mean(dataset_3$lv_std, na.rm = TRUE)) / sd(dataset_3$lv_std, na.rm = TRUE)
neigh_n_h <- c(0, 2, 4, 8, 12)
epreds_3 <- get_epreds(exp_3_m1, dataset_3, neigh_n_h = neigh_n_h, lv = lv) 
epreds_3$neigh_n_h_std <- factor(epreds_3$neigh_n_h_std, levels = unique(epreds_3$neigh_n_h_std), labels = paste0(neigh_n_h, " neighbours"), ordered = TRUE)

@@

# References {.unnumbered}

::: {#refs}
:::

## Appendix 1: Model diagnostics {#apx-diagnostics}

One way to diagnose the behaviour of Hamiltonian Monte Carlot (HMC, the algorithm used by Stan to explore the posterior distribution of a model) is to check whether the MCMC chains have converged. @fig-appendix-diagnostics shows the values sampled by the MCMC chains of each of the fixed coefficients of each model reported in the manuscript.  Evidence of chain convergence is provided by the same region of values being sampled across the final interations of the chain, as it is the case for the three models depicted.

```{r}
#| label: apxfig-appendix-diagnostics
#| fig-cap: "Traceplots of fixed regression coefficients of models in Experiments 1, 2, and 3."
#| fig-width: 9
#| fig-height: 9
var_labels <- c(
  "b_Intercept" = "Intercept",
  "b_freq_zipf_2_std" = "Frequency",
  "b_lv_std" = "Cognateness",
  "b_neigh_n_h_std" = "CLPN",
  "b_neigh_n_h_std:lv_std" = "Cognateness × CLPN",
  "b_group1" = "cat-ENG vs. spa-ENG"
)

list(exp_1_m0, exp_2_m0, exp_3_m1) |>
  set_names(paste0("Experiment ", 1:3)) |>
  map_dfr(gather_draws, `b_.*`, regex = TRUE, .id = "exp") |>
  dplyr::filter(.variable != "b_Intercept") |>
  mutate(
    .value = .value / 4,
    .variable = factor(.variable,
      levels = names(var_labels),
      labels = var_labels
    )
  ) |>
  ggplot(aes(.iteration, .value, colour = factor(.chain))) +
  facet_grid(.variable ~ exp, scales = "free") +
  geom_line() +
  labs(
    x = "Iteration",
    y = "Value",
    colour = "Chain",
    fill = "Chain",
    linetype = "Chain"
  ) +
  scale_colour_brewer(palette = "Reds") +
  theme(
    legend.position = "top",
    legend.box = "horizontal",
    panel.border = element_rect(
      fill = NA,
      colour = "grey"
    ),
    legend.justification = c(1, 0),
    axis.title.x = element_blank(),
    axis.title.y = element_blank()
  )
```

## Appendix 2: Pooled analyses of Experiments 1 and 3 {#apx-pooled}

Across Experiments 1 and 3, we found strong evidence that participants efficiently exploited phonological similarity to provide accurate translations for words in an unfamiliar language, provided that few phonological neighbours of higher lexical frequency were present. @fig-coefs summarizes the posterior distribution of the regression coefficients of the models in Experiments 1 to 3.

```{r}
#| label: apxfig-coefs
#| fig-cap: "Regression coefficients across Experiments 1, 2, and 3."
#| fig-width: 9
#| fig-height: 5
var_labels <- c(
  "b_Intercept" = "Intercept",
  "b_freq_zipf_2_std" = "Frequency",
  "b_lv_std" = "Similarity",
  "b_neigh_n_h_std" = "CLPN",
  "b_neigh_n_h_std:lv_std" = "Similarity × CLPN",
  "b_group1" = "cat-ENG vs. spa-ENG"
)

list(exp_1_m0, exp_2_m0, exp_3_m1) |>
  set_names(paste0("Experiment ", 1:3)) |>
  map_dfr(gather_draws, `b_.*`, regex = TRUE, .id = "exp") |>
  dplyr::filter(.variable != "b_Intercept") |>
  mutate(
    .value = .value / 4,
    .variable = factor(.variable,
      levels = names(var_labels),
      labels = var_labels
    )
  ) |>
  ggplot(aes(.value, reorder(exp, desc(exp)))) +
  facet_wrap(~.variable, scales = "free_x") +
  annotate(
    geom = "rect",
    xmin = -0.1 / 4,
    xmax = 0.1 / 4,
    ymin = -Inf,
    ymax = Inf,
    fill = "grey",
    alpha = 3 / 4,
    colour = NA
  ) +
  geom_vline(
    xintercept = 0,
    colour = "grey",
    linewidth = 3 / 4
  ) +
  stat_slab(aes(fill = after_stat(level)),
    point_interval = mode_hdi,
    .width = c(0.95, 0.89, 0.67, 1),
    outline_bars = TRUE,
    colour = "white",
    linewidth = 0.1,
    breaks = 20,
    normalize = "panels"
  ) +
  stat_spike(
    at = "median",
    size = 1,
    aes(linetype = "Median"),
    normalize = "panels"
  ) +
  labs(
    x = "Estimate", y = "Experiment",
    colour = "CrI",
    fill = "CrI",
    linetype = ""
  ) +
  scale_fill_brewer(palette = "Reds") +
  theme(
    legend.position = c(1, 0),
    legend.box = "horizontal",
    legend.justification = c(1, 0),
    axis.title.x = element_blank(),
    axis.title.y = element_blank()
  )
```
